Layered Grammar for Alluvial Plots

### 2020-01-20 paper.rmd
small paper revisions

Parallel sets plots might be viewed as a special subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis. These graphical constraints correspond to combinatorial constraints on the data, which amount to a id--key--value structure in which every id--key pair takes exactly one value (possibly 0 or `NA`).
In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]

The plots produced still vary---in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
The exceptional geoms of ggparallel offer common-angle as well as linear ribbons. Those of alluvial, ggforce, ggalluvial, and ggpcp offer a one-dimensional continuum between straight and x-spline ribbons.
The stats vertically arrange the elements (boxes and ribbons) at each axis.
These distinct elements are rendered by separate layers in ggforce, ggalluvial, and ggpcp, following the additive (`+`) syntax of ggplot2.
ggalluvial provides more levers of control over these statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]

@@

The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] with the geological terminology itemized above.
These alluvial plots are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them. The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes. While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots. (ggparallel and ggalluvial alone _only_ produce alluvial plots.)
This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
It is for this reason that i believe the typology and terminology are warranted.

@@

**Signed categorical data:**
@Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery. The total number of owners changed from year to year as exiters were not exactly matched by new entrants. In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot, reflected across the dimensional axis. Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots"). Alluvial plots provide a way to track cases and cohorts across such graphics, even if cases change sign. Future use cases may demonstrate the practical utility of this functionality.

>>>>

Parallel sets plots might be viewed as a subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis. These graphical constraints correspond to combinatorial constraints on the data, which amount to an id--key--value structure in which every id--key pair takes exactly one value (possibly zero or missing).
In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]

The plots produced still vary---in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
The exceptional geoms of ggparallel each offer common-angle as well as linear ribbons. Those of alluvial, ggforce, ggalluvial, and ggpcp offer a one-dimensional continuum between straight and x-spline ribbons.
The stats vertically arrange the elements (boxes and ribbons) at each axis.
These distinct elements are rendered by separate layers in ggforce, ggalluvial, and ggpcp, following the additive (`+`) syntax of ggplot2.
ggalluvial provides more levers of control over these statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]

@@

The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces,[^plots] with the geological terminology introduced above.
These alluvial plots are distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a real-valued plotting dimension perpendicular to that of flow, along which these elements are stacked, so that gaps between them are precluded. The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could arrange boxes corresponding to repeated categorical decompositions differently at different axes. While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots. (ggparallel and ggalluvial alone _only_ produce alluvial plots.)
This feature is particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
It is for this reason that i believe the typology and terminology are warranted.

@@

**Signed categorical data:**
@Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery. The total number of owners changed from year to year as exiters were not exactly matched by new entrants. In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot, reflected across the dimensional axis.[^negative] Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots"). Alluvial plots provide a way to track cases and cohorts across such graphics, even if cases change sign. Future use cases may demonstrate additional practical uses for this functionality.

[^negative]: The authors should be credited with this innovation, which i only implemented in ggalluvial after learning about their workaround to create it using a previous version.
